---
title: "Sample Movie Catalog"
format:
  html:
    theme: cosmo
    code-tools: true
    code-fold: true
    df-print: paged
    toc: true
    toc-depth: 2
    toc-expand: 3
editor: source
---

*This notebook demonstrates how you can use Quarto Markdown to generate a report of records from a table. Specifically, we will create a catalog of ~100 movie titles with plot summaries, and use Quarto's built-in TOC feature to create a TOC pane from level 1 & 2 headings.*

A rendered version of this notebook can be found at: <https://ucanr-igis.github.io/sketchbook/movie-catalog.html>

# Setup

Load packages and import an Excel file with plot summaries of 1000 movies.

```{r}
library(readxl)

movies_tbl <- here::here("data/movie-sampler.xlsx") |> 
  readxl::read_xlsx()

#dim(movies_tbl)
```

## Preview the data

The first six rows of the table we just imported look like:

```{r}
head(movies_tbl)
```

\

# Movie Summaries

The code chunk below creates an entry for each film.

```{r results='asis'}
for (i in 1:100) {
  cat("## ", movies_tbl[i,"MovieTitle", drop=TRUE], "\n\n")
  cat("**Running time:** ", movies_tbl[i,"RunningTime", drop=TRUE], "\n\n")
  cat("**Year:** ", movies_tbl[i,"Year", drop=TRUE], "\n\n")
  cat("**Plot Summary:** ", movies_tbl[i,"PlotSummary", drop=TRUE], "\n\n")
}
```



